An efficient acyclic contact planner for multiped robots

### 2018-02-19 reviews/reviews_tro/reviewer_response.tex

\quot{* Finally, I would suggest providing explicit git SHAs or tags for all
of the github links. I've personally made the mistake of linking to a
github project in a paper, only to have that project change and
eventually delete the specific code I was citing. }
\as{The Reviewer is absolutely right. We indeed encounter the issue with other researchers that are currently using our planner. 
To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project, on a separate branch
dedicated to the installation of the planner.
Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the Reviewer.}
\qt{(Section VII) Both HPP and our planner can be simply downloaded and compiled by following the instructions on
\url{https://humanoid-path-planner.github.io/hpp-doc/download.html?branch=rbprm}. \\
(Appendix C) To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software. }
\done

\section{Diff files between the original and current submission}
We now join a diff file comparing the original and current submissions.
The text appearing in blue denotes the editions. Most of the Figures have been removed, with the exception of the new Figure 8, to fit the size limit for the answer letter. We think that this diff file will help the Reviewers locate precisely
where the modifications occured in the document. The Reviewers will also notice that we reformulated the description our P3 solver in Appendix C, to better clarify what variables
are used, and what variables are implicit (that is computed from the others).

### 2018-02-19 final.tex
16 pages with bio

%~ \setcounter{secnumdepth}{3}
\usepackage{moreverb,url}
\usepackage{makeidx}         % allows index generation
\usepackage[percent]{overpic}
\usepackage{array} 
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{listings}
\usepackage{color}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[Symbol]{upgreek}

@@

The reduction breaks the algorithm complexity encountered
in previous works, resulting in the first interactive
implementation of a contact planner (open source). While
no contact planner has yet been proposed with theoretical
completeness, we empirically show the interest of our framework:
in a
few seconds, with high success rates, we generate complex contact plans for various scenarios and two robots, HRP-2 and HyQ. These plans are validated either in dynamic simulations, or on the real HRP-2 robot.

@@

Once the contact plan is known, efficient approaches exist to generate a dynamically feasible motion~\citep{Carpentier2016}.
 %~ However in general
%~ the contact plan is manually designed by a time consuming trial-and-error procedure. Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in cluttered, human-centered environments, which comprise constraining obstacles (stairs...).
In the specific (and simple) case of gaited biped locomotion on flat ground, 
%~ efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location. Furthermore, 
choosing the effector with which to create a contact is trivial because walking follows a cyclic pattern (the left foot always follows the right foot).
Efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
In the general case planning complex contact interactions is extremely challenging:
At any given time a contact must be chosen between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location). Furthermore, a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles; we say that the contact manifold is foliated~\citep{simeon-manipulation-04}. 
%~ Lastly, the contact manifold has a null measure and is thus impossible to sample. This restriction prevents the use of efficient sampling-based planners.
The
foliation prevents the use of efficient sampling-based
planners
for two reasons. (i) First, each sub-manifold of the         
foliation
has
a zero measure and cannot be directly sampled.         
A
sample
is
rather obtained by sampling a “free flying”         
configuration
and
explicitly projecting it in contact (which is a         
costly
numerical
operation).
(ii) Second, the foliated topology       
turns
the
exploration
by
spreading
a graph of configurations        
(probabilistic
road-map,
rapidly
exploring
random trees) into      
an
inefficient
random
process,
where
many
useless nodes       
are
sampled
on
parallel
sub-manifolds.
The
total
algorithmic       
complexity
of
classical
contact
planners
comes
from
both the        
number
of
graph
nodes
sampled
during
exploration
(ii) and
the cost of the projection when sampling new configurations
(i).

@@

Additionally to robotics, acyclic motion planning is also a problem of interest in neurosciences, biomechanics, and virtual character animation.
Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}. These approaches are by definition not able to discover complex behaviors in unforeseen contexts.

@@

%~ because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. 
Compared to previous approaches, our solution has two main novelties:

@@

\begin{figure}[t]
\centering
  \begin{overpic}[width=0.8\linewidth]{figures/reachability_cond}
	\end{overpic}
\caption{The reachability condition is verified by the right configuration: the trunk (red) is free of collisions while the limbs reachable workspace (green) intersect the environment.}
		   \label{fig:reach_cond}
\end{figure}

@@

At each step of Algorithm~1, \textsc{GenFullBody} (Algorithm~\ref{alg:pi}) is called with the previous state as a parameter, as well
as a new root configuration. \textsc{GenFullBody} returns a new contact configuration, if it succeeded
in computing a configuration with only one contact switch occurring~;
otherwise, the method \textsc{IntermediateContactState} (Algorithm~\ref{alg:repo}) is called.
It repositions an end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
This repositioning allows to increase the odds that the contact can be maintained at the next step.
The method \textit{Length} returns the length of a list~; \textit{LastElement} returns the last element
of a list~; \textit{Element} returns the element contained by a list at a given index~; and \textit{PushBack} inserts an element at the end of a list.
%~ Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.

@@

In Alg.~\ref{alg:pi}, the procedure \textit{MaintainContact}$(pState,\mathbf{q}^0,k)$ performs inverse kinematics to reach the previous contact position for the limb.
If it succeeds, the new limb configuration is assigned to $k$. If it fails, a random collision free configuration is assigned to $k$.
The method \textit{IsInStaticEquilibrium} returns whether a given state is in static equilibrium. \textit{CreateState} creates a new state, given
a root configuration and the list of active contacts.
$ContactLimbs$ (respectively $FreeLimbs$) returns the list of limbs that are in contact (respectively \textit{not} in contact).  \textit{MarkContact} (respectively \textit{MarkFree})  marks a limb as in contact (respectively \textit{not} in contact). These methods follow a FIFO approach: the first limb chronologically marked as in contact (respectively not in contact) is returned first by $ContactLimbs$ (respectively $FreeLimbs$). This allows the algorithm to be deterministic even though it can handle acyclic motions.
%~ $add\_contact\_limb(newState,k)$ marks a limb as in contact at a given state.
%
%~ The pseudo code for the method \textsc{IntermediateContactState} is given by Algorithm~\ref{alg:repo}.
%
\textit{GenerateContact}$(state,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
It generates a contact configuration in static equilibrium, and assigns the corresponding configuration of the limb $k$ to the state configuration.
If it fails, $k$ remains unchanged if it is collision free, otherwise it is assigned a random collision free configuration.

@@

According to Coulomb's law, the non-slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
As classically done, we linearize the friction cone in a conservative fashion with a pyramid included in it, described by four generating rays of unit length. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{\boldsymbol\upgamma}_{i1} & \mathbf{n}_{i} -\mu \mathbf{\boldsymbol\upgamma}_{i1} & \mathbf{n}_{i} + \mu \mathbf{\boldsymbol\upgamma}_{i2} & \mathbf{n}_{i} - \mu \mathbf{\boldsymbol\upgamma}_{i2}}^T
\end{equation*}

@@

From the Newton-Euler equations, to be in static equilibrium the contact forces have to compensate the gravitational forces:

@@

\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $20$.

@@

\noindent\textbf{Heuristics:} $h_w$ for all legs. The robustness threshold $b_0$ is set to $10$.

@@

\subsubsection{HRP-2 -- Path re-planning (Figure~\ref{fig:re-planning})}
In this long scene, HRP-2 plans a path through several obstacles. The scene is edited during the execution of the motion: a stair is added,
some stepping stones are removed, and part of the final staircase is deleted. All these modifications require re-planning.

@@

\subsubsection{Choosing the heuristics} \label{sec:heuristichoices}
In our conference paper~\citep{tonneauisrr15}, the computed motions were generated using the EFORT heuristic.
EFORT is designed for tasks requiring large magnitude contact forces (such as pushing / pulling / climbing). 
In locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
configurations close to singularities (and joint limits). While this did not significantly impact
the generation of the plan, the resulting interpolation turned out to be harder.
For this reason, we prefer to use our manipulability-based heuristic for the legs of the robot, but we still
use EFORT for the arms, which results in fewer contact repositionings.

@@

\subsubsection{Discretization of the guide path} \label{sec:disc}
The discretization step is a user-defined, fixed parameter. The step
has an influence on the output of the planner: if too large steps are taken,
the planner may fail since we impose the constraint that only one contact change might occur
between two consecutive steps. On the other hand, a small step will not impact the success rate of the planner, 
but may generate unnecessary states. In most scenarios the torso of HRP-2 moves about 15 cm between two postures, but only 3 cm
for the car egress scenario to handle the geometry of the car.
For future work, we would like to automatically adapt the size of the discretization step to the complexity of the environment.

@@

For HRP-2, most of the time was spent performing inverse kinematics.
This is not surprising considering the number of calls to the methods: IK projection is used intensively to maintain contact continuity between two postures; 
it is also applied every time a new candidate needs to be evaluated. In particular for the car egress scenario,
the kinematic constraints are very demanding to avoid collisions.

@@

\subsection{Step 2: computing the torso scaling workspace $W^0$ of the robot}
To define the volume $W^0$ of HRP-2, we proceed in an empirical manner.
First, we compute the bounding boxes of the robot torso, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
Then, we perform a scaling of these boxes by a factor $s$. 
The higher $s$ is, the more likely sampled configurations are to be feasible, but the less complete is the approach.
To compute the appropriate value of $s$, we proceed as described in Section~\ref{sec:params}, and choose empirically
$s^*=1.2$ as the appropriate value for HRP-2.
%~ \section{Heuristics and criteria for contact selection}
%~ \label{sec:heuristics}
% !TEX root =  ../main.tex
%~ \subsection{Manipulability-based heuristics for contact selection}
\section{Manipulability-based heuristics for contact selection}
\label{sec:heuristics}
This Section proposes two heuristics to select a contact that optimizes desired capabilities.
%~ These heuristics allow to locally optimize a contact location with respect to a particular task.
For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion, or to stay away from singular configurations.
We derive these heuristics from the work by \cite{Yoshikawa1984}, recalled here. %, that we present first.

@@

\begin{figure}[!tbp]
  \centering
	\begin{overpic}[width=1\linewidth]{figures/ellipsoid}
		\put (8.5,1.2) {\small{Velocity ellipsoid}}
		\put (62.5,1.2) {\small{Force ellipsoid} \tiny{(scale 0.5)}}
	\end{overpic}
  \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 DoFs and 2 segments.
Only the horizontal and vertical speeds are shown (not the rotation speeds).}
 %~ since it would require being able to draw in four dimensions.}
		   \label{sec:efort_ellipsoid}
\end{figure}

We consider: a limb configuration $\mathbf{q}^k$; its end effector position $\mathbf{p}^k$; its Jacobian matrix
$\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
Yoshikawa~\cite{Yoshikawa1984} defines the velocity(~\ref{eq:vel}) and force~(\ref{eq:for}) ellipsoids:

@@

They describe the set of end-effector velocities (respectively forces) that can
be reached under the constraint $||\dot{\mathbf{q}}||^2 \leq 1$ for the current configuration.
The longer the axis of the ellipsoid is, the more important the velocity (resp. force) of the end-effector the direction of the axis can be (Figure~\ref{sec:efort_ellipsoid}).

\subsubsection{Manipulability-based heuristics}
From these definitions, we can derive two useful heuristics, that all account for the environment and the task being performed.
The first one, EFORT, was introduced by \cite{Tonneau2014}; the second one derives the manipulability measure proposed by Yoshikawa~\cite{Yoshikawa1984}.

With EFORT, we define the efficiency of a configuration as the ability of a limb to exert a force in a given direction.
We thus consider the force ellipsoid as a basis for our heuristic.
In a given direction $\uprho$, the length of the ellipsoid is given by the force-transmission ratio \citep{1087795}:
\begin{equation*}
f_\mathsf{T}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}	^{T})\uprho]^{-\frac{1}{2}}
\end{equation*}

In our problem, to compare candidate configurations, we include the quality of the contact surface, and choose $\uprho$ as the direction
opposite to the local motion (given by the difference between two consecutive root positions):

@@

%~ If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
%~ \begin{equation}
%~ h_{vel}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\uprho]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \uprho)
%~ \end{equation}

%~ $h_{\textrm{\it EFORT}}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
$h_{\textrm{\it EFORT}}$ will favor contacts that allow large efforts.
$EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
In other less demanding cases, manipulability can also be considered to avoid singularities.
To do so, we can consider the manipulability measure $h_{w}$, also given by Yoshikawa~\cite{Yoshikawa1984}:

\begin{equation} \label{ellipsoid}
h_{w}(\mathbf{q}) = \sqrt{det(\mathbf{J}\mathbf{J}^T)}
\end{equation}
$h_{w}$ measures the ``distance'' of a given configuration to singularity. When $h_{w}$ is equal to 0, the configuration is singular;
the greater $h_{w}$ is, the further away the configuration is from singularity.

%~ \input{body/equilibrium}
% !TEX root =  ../main.tex
\section{Addressing $\mathcal{P}_3$}
\label{app:optim}
The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
%~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
To address the other scenarios, we propose a new implementation, entirely open source (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#optim} ), which can be integrated directly in our motion planner software. This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
The center-of-mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.

@@

\noindent where $\dot{L}$ is the angular momentum expressed at the com.
Eq.~\ref{eq:new_eul_acc} defines a 6-dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
this cone determines all the admissible wrenches $\mathbf{w}$ that can be generated by contact forces inside their friction cones.
The face form of $\mathcal{K}$ can be computed using the double description method~\citep{Fukuda1996}, resulting in the following linear inequalities:

@@

%~ \begin{IEEEbiographynophoto}{Steve Tonneau}
%~ TODO
%~ \end{IEEEbiographynophoto}

>>>>

%~ \setcounter{secnumdepth}{3}
\usepackage{moreverb,url}
\usepackage{makeidx}         % allows index generation
\usepackage[percent]{overpic}
\usepackage{array} 
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{listings}
\usepackage{color}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage[Symbol]{upgreek}
\usepackage[utf8]{inputenc}

@@

The reduction breaks the algorithm complexity encountered
in previous works, resulting in the first interactive
implementation of a contact planner (open source). While
no contact planner has yet been proposed with theoretical
completeness, we empirically show the interest of our framework:
in a
few seconds, with high success rates, we generate complex contact plans for various scenarios and two robots, HRP-2 and HyQ. These plans are validated in dynamic simulations or on the real HRP-2 robot.

@@

Once the contact plan is known, efficient approaches exist to generate a dynamically feasible motion~\citep{Carpentier2016}.
 %~ However in general
%~ the contact plan is manually designed by a time consuming trial-and-error procedure. Automatically addressing the acyclic contact planning problem is thus a pre-requisite to the deployment of robots in cluttered, human-centered environments, which comprise constraining obstacles (stairs...).
In the specific (and simple) case of gaited biped locomotion on flat ground, 
%~ efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location. Furthermore, 
choosing the effector with which to create a contact is trivial because walking follows a cyclic pattern (the left foot always follows the right foot).
Efficient tools such as the capture point~\citep{Pratt2006} can be used to compute the next contact location.
In the general case planning complex contact interactions is extremely challenging:
At any given time a contact must be chosen between infinitely many possibilities (often a combinatorial discrete choice for the effector and contact surface, and a continuous choice for the contact location). Furthermore, a contact choice constrains kinematically and dynamically the possible motions, and there is no analytical way to verify whether this choice brings the robot one step closer to the desired goal or to a dead end, especially in the presence of obstacles; we say that the contact manifold is foliated~\citep{simeon-manipulation-04}. 
%~ Lastly, the contact manifold has a null measure and is thus impossible to sample. This restriction prevents the use of efficient sampling-based planners.
The
foliation prevents the use of efficient sampling-based
planners
for two reasons. (i) Each sub-manifold of the         
foliation
has
a zero measure and cannot be directly sampled.         
A
sample
is
rather obtained by sampling a “free flying”         
configuration
and
explicitly projecting it in contact (which is a         
costly
numerical
operation).
(ii) The foliated topology       
turns
the
exploration
by
spreading
a graph of configurations        
(probabilistic
road-map,
rapidly
exploring
random trees) into      
an
inefficient
random
process,
where
many
useless nodes       
are
sampled
on
parallel
sub-manifolds.
The
total
algorithmic       
complexity
of
classical
contact
planners
comes
from
both the        
number
of
graph
nodes
sampled
during
exploration
(ii) and
the cost of the projection when sampling new configurations
(i).

@@

Additionally to robotics, acyclic motion planning is also a problem of interest in biomechanics and virtual character animation.
Early contributions in the latter field rely on local adaptation of motion graphs \citep{citeulike:220163}, or ad-hoc construction of locomotion controllers \citep{Pettre:2003:LPD:846276.846313}. These approaches are by definition not able to discover complex behaviors in unforeseen contexts.

@@

%~ because we believe it is the most promising direction~\citep{DBLP:conf/iser/EscandeKMG08} to break the complexity of acyclic contact planning. 
Our solution presents two main novelties:

@@

\begin{figure}[t]
\centering
  \begin{overpic}[width=0.8\linewidth]{figures/reachability_cond}
	\end{overpic}
\caption{The reachability condition is verified if the trunk (red) is collision-free while the limbs reachable workspace (green) intersect the environment.}
		   \label{fig:reach_cond}
\end{figure}

@@

At each step of Algorithm~1, \textsc{GenFullBody} (Algorithm~\ref{alg:pi}) is called with the previous state as a parameter, as well
as a new root configuration. \textsc{GenFullBody} returns a new contact configuration, if it succeeded
in computing a configuration with only one contact switch occurring~;
otherwise, the method \textsc{IntermediateContactState} (Algorithm~\ref{alg:repo}) is called.
It repositions an end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
This repositioning allows to increase the odds that the contact can be maintained at the next step.
The method \textit{Length} returns the length of a list~; \textit{LastElement} returns the last element
of a list~; \textit{Element} returns the element of a list at a given index~; and \textit{PushBack} inserts an element at the end of a list.
%~ Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.

@@

In Alg.~\ref{alg:pi}, the procedure \textit{MaintainContact}$(pState,\mathbf{q}^0,k)$ performs inverse kinematics to reach the previous contact position for the limb.
If it succeeds, the new limb configuration is assigned to $k$. If it fails, a random collision free configuration is assigned to $k$.
The method \textit{IsInStaticEquilibrium} returns whether a given state is in static equilibrium. \textit{CreateState} creates a new state, given
a root configuration and the list of active contacts.
$ContactLimbs$ (respectively $FreeLimbs$) returns the list of limbs that are in contact (respectively \textit{not} in contact).  \textit{MarkContact} (respectively \textit{MarkFree})  marks a limb as in contact (respectively \textit{not} in contact). These methods follow a FIFO approach: the first limb chronologically marked as in contact (respectively not in contact) is returned first by $ContactLimbs$ (respectively $FreeLimbs$). This allows the algorithm to be deterministic even though it can handle acyclic motions.
%~ $add\_contact\_limb(newState,k)$ marks a limb as in contact at a given state.
%
%~ The pseudo code for the method \textsc{IntermediateContactState} is given by Algorithm~\ref{alg:repo}.
%
\textit{GenerateContact}$(state,k)$ is a call to the contact generator presented in the following Section~\ref{sec:single_contact}.
It generates a contact configuration in static equilibrium, and assigns the corresponding configuration of the limb $k$ to the state configuration.
If it fails, $k$ remains unchanged if collision free, else it is assigned a random collision free configuration.

@@

According to Coulomb's law, the non-slipping condition is verified if all the contact forces lie in the friction cone defined by the surface.
Classically, we linearize the friction cone in a conservative fashion with a pyramid included in it, described by four generating rays of unit length. We choose for instance:
\begin{equation*}
\mathbf{V}_{i} = \mat{\mathbf{n}_{i} + \mu \mathbf{\boldsymbol\upgamma}_{i1} & \mathbf{n}_{i} -\mu \mathbf{\boldsymbol\upgamma}_{i1} & \mathbf{n}_{i} + \mu \mathbf{\boldsymbol\upgamma}_{i2} & \mathbf{n}_{i} - \mu \mathbf{\boldsymbol\upgamma}_{i2}}^T
\end{equation*}

@@

From the Newton-Euler equations, to be in static equilibrium the contact forces have to compensate the gravity:

@@

\noindent\textbf{Heuristics:} $h_w$. The robustness threshold $b_0$ is set to $20$.

@@

\noindent\textbf{Heuristics:} $h_w$. The robustness threshold $b_0$ is set to $10$.

@@

\subsubsection{HRP-2 -- Path re-planning (Figure~\ref{fig:re-planning})}
In this long scene, HRP-2 plans a path through several obstacles. The scene is edited during the execution of the motion: a stair is added,
stepping stones are removed, as for parts of the final staircase. All these modifications require re-planning.

@@

\subsubsection{Choosing the heuristics} \label{sec:heuristichoices}
In our conference paper~\citep{tonneauisrr15}, the computed motions were generated using the EFORT heuristic.
EFORT is designed for tasks requiring large magnitude contact forces (such as pushing / pulling / climbing). 
In locomotion tasks, such as the stair scenario, one issue with EFORT is that it tends to generate
configurations close to singularities (and joint limits). While this did not significantly impact
the generation of the plan, the resulting interpolation turned out to be harder.
For this reason, we use our manipulability-based heuristic for the legs, but still
use EFORT for the arms, which results in fewer contact repositionings.

@@

\subsubsection{Discretization of the guide path} \label{sec:disc}
The discretization step is a user-defined, fixed parameter. The step
has an influence on the output of the planner: if too large steps are taken,
the planner may fail since we impose the constraint that only one contact change might occur
between two consecutive steps. On the other hand, a small step will not impact the success rate of the planner, 
but may generate unnecessary states. In most scenarios the torso of HRP-2 moves about 15 cm between two postures, but only 3 cm
for the car egress scenario to handle the geometry of the car.
For future work, we would like to automatically adapt the size of the discretization step.

@@

For HRP-2, most of the time was spent performing inverse kinematics.
This is not surprising considering the number of calls to the methods: IK projection is used intensively to maintain contact continuity between two postures; 
it is also applied every time a new candidate needs to be evaluated. In particular for the car egress scenario,
the collision avoidance constraints are demanding.

@@

\subsection{Step 2: computing the torso scaling workspace $W^0$}
To define the volume $W^0$ of HRP-2, we proceed in an empirical manner.
First, we compute the bounding boxes of the robot torso, head, and upper legs (Figure~\ref{fig:hrp2_w} -- red shapes).
Then, we perform a scaling of these boxes by a factor $s$. 
The higher $s$ is, the more likely sampled configurations are to be feasible, but the less complete is the approach.
To compute the appropriate value of $s$, we proceed as described in Section~\ref{sec:params}, and choose empirically
$s^*=1.2$ as the appropriate value for HRP-2.
%~ \section{Heuristics and criteria for contact selection}
%~ \label{sec:heuristics}
% !TEX root =  ../main.tex
%~ \subsection{Manipulability-based heuristics for contact selection}
\section{Manipulability-based heuristics for contact selection}
\label{sec:heuristics}
This Section proposes two heuristics to select a contact that optimizes desired capabilities.
%~ These heuristics allow to locally optimize a contact location with respect to a particular task.
For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion, or to stay away from singular configurations.
We derive these heuristics from the work by \cite{Yoshikawa1984}, recalled here. %, that we present first.

@@

We consider: a limb configuration $\mathbf{q}^k$; its end effector position $\mathbf{p}^k$; its Jacobian matrix
$\mathbf{J}^k(\mathbf{q}^k)$; a force $\mathbf{f}$ exerted by the end effector. For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.
Yoshikawa~\cite{Yoshikawa1984} defines the velocity (\ref{eq:vel}) and force~(\ref{eq:for}) ellipsoids:

@@

They describe the set of end-effector velocities (respectively forces) that can
be reached under the constraint $||\dot{\mathbf{q}}||^2 \leq 1$ for the current configuration.
The longer the axis of the ellipsoid is, the more important the velocity (resp. force) of the end-effector the direction of the axis can be.
%~ (Figure~\ref{sec:efort_ellipsoid}).

\subsubsection{Manipulability-based heuristics}
From these definitions, we derive two heuristics that account for the environment and the task being performed.
The first one, EFORT, was introduced by \cite{Tonneau2014}; the second one derives the manipulability measure proposed by Yoshikawa~\cite{Yoshikawa1984}.

With EFORT, we define the efficiency of a configuration as the ability of a limb to exert a force in a given direction.
In a given direction $\uprho$, the length of the ellipsoid is given by the force-transmission ratio \citep{1087795}:
\begin{equation*}
f_\mathsf{T}(\mathbf{q}, \uprho) = [\uprho^{T}(\mathbf{J}\mathbf{J}	^{T})\uprho]^{-\frac{1}{2}}
\end{equation*}

To compare candidate configurations, we include the quality of the contact surface, and choose $\uprho$ as the direction opposite to the local motion (given by the difference between two consecutive root positions):

@@

$h_{\textrm{\it EFORT}}$ will favor contacts that allow large efforts. Alternatively, the manipulability measure can be considered~\cite{Yoshikawa1984}:

\begin{equation} \label{ellipsoid}
h_{w}(\mathbf{q}) = \sqrt{det(\mathbf{J}\mathbf{J}^T)}
\end{equation}
%~ $h_{w}$ measures the ``distance'' of a given configuration to singularity. When $h_{w}$ is equal to 0, the configuration is singular;
%~ the greater $h_{w}$ is, the further away the configuration is from singularity.

%~ \input{body/equilibrium}
% !TEX root =  ../main.tex
\section{Addressing $\mathcal{P}_3$}
\label{app:optim}
The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
%~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
To address the other scenarios, we propose a new implementation, entirely open source (\url{http://stevetonneau.fr/files/publications/isrr15/tro_install.html\#optim} ), which can be integrated directly in our motion planner software. This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
The obtained COM trajectory is then turned into a collision-free whole-body trajectory.

@@

\noindent where $\dot{L}$ is the angular momentum expressed at the com.
Eq.~\ref{eq:new_eul_acc} defines a 6-dimensional cone $\mathcal{K}$ ~\citep{qiu:dhm:2011,Caron2015}. For a given set of contacts,
this cone determines all the admissible wrenches $\mathbf{w}$ that can be generated by contact forces inside their friction cones.
The face form of $\mathcal{K}$ can be computed using the double description method~\citep{Fukuda1996}:

@@

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/steve}}]{Steve Tonneau} is a post-doctorate in the Gepetto team at LAAS-CNRS. He received the joint M.Eng. and M.Sc. degrees from Insa Rennes in 2008. After 4 years in the industry he defended his Phd in 2015 at the INRIA/IRISA Mimetic research team. His research focuses on motion planning based for legged avatars in arbitrary situations. Applications include computer graphics animation and legged robotics.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/andrea}}]{Andrea Del Prete} received his degree in Computer
Engineering (with honors) from the 2nd faculty of
the University of Bologna (Italy) in 2009. In March
2013 he got his PhD from the Cognitive Humanoids
laboratory, Italian Institute of Technology, Genova
(Italy). From 2014 to 2017 he had been a PostDoc
at LAAS-CNRS in Toulouse (France), working on
optimization-based control of the humanoid robot
HRP-2. Since 2018 he is a senior researcher at
the Max Planck Institute for Intelligent Systems, in
Tuebingen (Germany).
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/julien}}]{Julien Pettré}
is a computer scientist. He is junior researcher at Inria, the French National Institute for Research in Computer Science and Control, in the Lagadic team. He received PhD from the University of Toulouse III in 2003, and Habilitation from the University of Rennes I in 2015. From 2004 to 2006, he was postdoctoral fellow at EPFL in Switzerland. He joined Inria in 2006.
Julien Pettré is currently coordinator of the European H2020 Crowdbot project (2018-21), dedicated to the design of robot navigation techniques for crowded environments. He previously coordinated the national ANR JCJC Percolation project (2013-17) dedicated to the design of new microscopic crowd simulation algorithms, as well as the national ANR CONTINT Chrome project, dedicated to efficient and designer-friendly techniques for crowd animation.
His research interests are crowd simulation, computer animation, virtual reality, robot navigation and motion planning. 
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/chonhyon}}]{Chonhyon Park}
received the B.S. degree and the M.S. degree in computer science and engineering from the Seoul National University, Seoul, South Korea 2005 and 2007, respectively.

He received the Ph.D. degree in computer science at University of North Carolina at Chapel Hill in 2016.

His current research interests include motion and path planning,  navigation of virtual characters, and many-core computing.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/dinesh}}]{Dinesh Manocha}
is currently the Phi Delta
Theta/Mason Distinguished Professor of Computer
Science at the University of North Carolina at
Chapel Hill. He received his Ph.D. in Computer
Science at the University of California at Berkeley
1992. Along with his students, Manocha has
also received 15 best paper awards at the leading
conferences. He has published more than 480
papers and some of the software systems related
to collision detection, GPU-based algorithms and
geometric computing developed by his group have
been downloaded by more than 200,000 users and are widely used in the
industry. He has supervised 33 Ph.D. dissertations and is a fellow of ACM, AAAI,
AAAS, and IEEE. He received Distinguished Alumni Award from Indian
Institute of Technology, Delhi.
\end{IEEEbiography}

\begin{IEEEbiography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{figures/bio/nicolas}}]{Nicolas Mansard}
received the joint M.Eng. and
M.Sc. degrees in 2003, robotics and image process-
ing from Nationale Supérieure d'Informatique et de
Mathématiques Appliquées de Grenoble,
France, and University Joseph Fourier, Grenoble,
France, respectively, and the Ph.D. degree in robotics
for his work with Lagadic Group, Institut National
de Recherche en Informatique et en Automatique,
Rennes, France, in 2006. He spent one year with
Stanford University, Stanford, CA, with O. Khatib
and one year with JRL-Japan with A. Kheddar. He
is currently with the GEPETTO Group, LAAS- CNRS, Toulouse, France.
He was an Invited Researcher with Emo Todorovs Group, University of
Washington, in 2013. His research interest includes the generation of dynamic
sensor-based robot movements. Dr. N. Mansard received the CNRS Bronze
medal in 2015.
\end{IEEEbiography}
